\textsc{HubbleCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-01-01 paper/main.tex
More text edits

\subsection{Data Selection and Pre-Processing}

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST) via the \texttt{astroquery} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. Note that these are not science-grade observations, but rather low-resolution images used for quick-look purposes; given the nature of associations we aim to learn, we deem this adequate for our current purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations. Images are resized to a maximum resolution per side of 512 pixels before saving. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, the abstract is excluded from the dataset.

@@

Raw abstracts summarize the corresponding successful HST observing proposal, which intends to make the case for allocating \hubble telescope time towards a particular set of observations. These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well. Although the abstracts can be used as-is as image captions, we explore the use of summarization via guided LLM generation to standardize the captions used for fine-tuning the CLIP model. The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the signal between text and images.

@@

As the base model, we use the CLIP-ViT-B/16 \citep{radford2021learning} variant trained by OpenAI. This model uses a (12-layer, 12-head, 768-embedding dimension) vision transformer as the image encoder and a (12-layer, 8-head, 512-embedding dimension) sequence transformer as the text backbone. Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension 512. The model was trained on 400 million image-text pairs from the internet.  % , and the CLIP family of models has shown strong performance on a variety of downstream tasks including zero-shot classification and image retrieval.

\subsection{Fine-tuning Procedure}
\label{sec:finetuning}

>>>>

\subsection{Data Selection and Pre-processing}

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST) via the \texttt{astroquery} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. Note that these are not science-grade observations, but rather low-resolution images used for quick-look purposes; given the nature of associations we aim to learn, we deem this adequate for our current purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations. Images are centered and resized to a resolution per side of 512 pixels before saving. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, the abstract is excluded from the dataset.

@@

Raw abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations. These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well. Although the abstracts can be used as-is as image captions, we explore the use of summarization via guided LLM generation to standardize the captions used for fine-tuning the CLIP model. The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the signal between text and images.

@@

As the base model, we use the CLIP-ViT-B/16 \citep{radford2021learning} variant trained by OpenAI. This model uses a (12-layer, 12-head, 768-embedding dimension) vision transformer as the image encoder and a (12-layer, 8-head, 512-embedding dimension) sequence transformer as the text backbone. The text encoder has a maximum length of 77-tokens and the image encoder a native resolution of $224\times224$. Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension 512. In total, the model has 149,620,737 trainable parameters. The model was trained on 400 million image-text pairs from the internet.

\subsection{Fine-tuning Procedure and Evaluation}
\label{sec:finetuning}

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized data or the raw proposal abstracts. When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum length of the text encoder. Images are downsampled to the native resolution of the image encoder and randomly rotated over training. Given the relatively modest size of the fine-tuning set, a batch size $|\mathcal B| = 32$ is used throughout. We note that positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract.

We explore three different methods of training the model on our domain dataset: \emph{(1)} Fine-tuning the entire network, starting from the pre-trained base model; \emph{(2)} Freezing the base image and text encoders, and training a small projection head on top of these; and \emph{(3)} Training the entire model from scratch.

The model is trained over 50,000 steps with 5000 linear warmup steps and cosine decay using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with peak learning rate of either $10^{-5}$ or $10^{-6}$ and weight decay $10^{-3}$. Training takes approximately 6 hours on 4 Nvidia A100 GPUs.
